---
title: "Estimating growth increment, productivity, and vital rates with ProdVital"
output: 
  html_document:
    toc: true
author: "John L. Godlee"
date: 2023-08-10
bibliography: ProdVital_vignette.bib
---

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.align = "center", linewidth = 60, cache = FALSE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

# Setup

Load the `{seosawr}` package, which contains some useful functions for cleaning
SEOSAW stem data. The `{seosawr}` package is located in the [SEOSAW Bitbucket
repository](https://bitbucket.org/miombo/seosaw). Refer to the `README.md` in
that repository for how to install the package. Load the `{dplyr}` package for
basic data manipulation. Load `{ProdVital}` which contains the core functions
for analysing phyto-demographic data.

```{r message=FALSE}
# Packages
library(dplyr)
library(seosawr)
library(ProdVital)
library(ggplot2)
```

Load the example dataset included in `{ProdVital}`:

```{r}
data(bicuar)
```

`bicuar` is a dataframe of stem data from two plots in Bicuar National Park,
southwest Angola, measured across three censuses. Censuses were conducted in
2019, 2021 and 2023. The original data is real, but it has been adjusted to
better illustrate the functionality of `{ProdVital}`. The data is in the SEOSAW
style, but with columns removed where they are not relevant to this workflow.
Each row is a stem measurement in a given census.

* `plot_id` - ID of the plot
* `census_date` - Census date of measurement 
* `stem_id` - ID value grouping measurements into stems
* `stem_status` - Vital status: alive, resprouting, dead
* `diam` - Stem diameter (cm)
* `agb` - Estimated aboveground biomass (Mg)

# Prepare the stem data

Stems may be missing from a particular census, despite being measured in
censuses before and after. Stem timelines should not have gaps, otherwise you
will get unrealistic estimates of recruitment, mortality, and biomass turnover
using the functions in `{ProdVital}`. We need to create these rows so we can
fill them with estimated values. We can do this using the
`seosawr::stemFillGen()` function:

```{r}
# Generate missing rows
plots <- unique(bicuar[,c("plot_id", "census_date")])
missing_records <- stemFillGen(bicuar, plots)

# All the new records should be in the middle census
stopifnot(all(missing_records$census_date == 2021))

# Add the missing rows to the stem data
bicuar_pad <- bind_rows(bicuar, missing_records)
```

Sometimes, a stem will appear to have died only to resurrect at a later date.
Stems might also be missing stem status values. We can try to fix these issues
using the `seosawr::statusImputGen()` function. 

```{r}
# Impute mortality status
status_est <- statusImputGen(bicuar_pad)

# Add new columns to stem data
bicuar_sfix <- bind_cols(bicuar_pad, status_est)
```

Now that the stem status values have been fixed, we should remove measurements
of dead stems. The functions in `{ProdVital}` assume that the stem data only
contain measurements from living stems. Measurements from dead stems should be
excluded from the stem data. The absence of a stem from a particular census
after it was initially measured implies that the stem was dead during that
census. Similarly the absence of a stem from a particular census before it was
initially measured implies that the stem had not yet recruited. It's up to you
whether you count resprouting stems (`stem_status` "r", stem dead above point
of measurement but with living material elsewhere on the tree) as alive or
dead:

```{r}
bicuar_alive <- bicuar_sfix %>%
  filter(stem_status_est == "a")
```

Stem diameter measurements might be missing. These values should be imputed if
you are calculating plot-level productivity, otherwise you will get unrealistic
estimates of growth and loss. A simple way to impute missing diameter values is
to take the mean of measured diameters before and after the missing diameter:

```{r}
bicuar_clean <- bicuar_alive %>% 
  group_by(plot_id, stem_id) %>% 
  mutate(
    diam = ifelse(is.na(diam), mean(diam, na.rm = TRUE), diam),
    agb = ifelse(is.na(agb), mean(agb, na.rm = TRUE), agb))

stopifnot(all(!is.na(bicuar_clean$diam)))
```

It is important to understand that this naive approach doesn't consider that
some diameter measurements may contain errors, what should happen if the
missing value is at the start or end of the stem timeline, or what should
happen if there are multiple consecutive missing values. In these cases, more
complex approaches will be required to impute missing diameter values.

If you only want to calculate stem growth increment, these missing values can
simply be removed:

```{r}
bicuar_excl <- bicuar_clean %>% 
  filter(!is.na(diam))
```

# Identify stem fate

Using the `obsID()` function, it is possible to identify the fate of individual
stems within a given census interval. __Recruits__ are stems which were not
present in the initial census, but are present in the final census.
__Survivors__ are stems which are present in both the initial and final census.
__Deaths__ are stems which are present in the initial census but not in the
final census:

```{r}
bicuar_clean$plot_stem <- paste(bicuar_clean$plot_id, 
  bicuar_clean$stem_id, sep = ":")

obs1 <- obsID(bicuar_clean, t0 = "2019", tT = "2021",
  type = c("rec", "sur", "mor"), group = "plot_stem",
  census = "census_date")

lapply(obs1, head)
```

Note that `plot_stem` was created to provide single ID column, but this is also
valid:

```{r}
obs2 <- obsID(bicuar_clean, t0 = "2019", tT = "2021",
  type = c("rec", "sur", "mor"), group = c("plot_id", "stem_id"),
  census = "census_date")

lapply(obs2, head)
```

`obsID()` serves as a building block in many other functions in `{ProdVital}`,
like `prodKohyama()`, `vitalKohyama()` and `obsSurGrowth()`.

# Observed stem growth, recruitment growth, and mortality loss

The `obsSurGrowth()` function can calculate the growth increment of all stems
which survived between two censuses. The argument `w` gives the column used to
calculate growth. The function returns a named vector of stem growth increments
between the two census dates. The names refer to the `group` ID column(s) used
to define individuals. 

```{r}
obsSurGrowth(bicuar_clean, t0 = "2019", tT = "2021", w = "diam", 
  group = c("plot_id", "stem_id"), census = "census_date")
```

With options `full = TRUE`, `obsSurGrowth()` returns a dataframe rather than a
named vector, which may be easier to join back to the original data using the
columns in `group`.

```{r}
obsSurGrowth(bicuar_clean, t0 = "2019", tT = "2021", w = "diam", 
  group = c("plot_id", "stem_id"), census = "census_date", full = TRUE)
```

Similarly, the `obsMorLoss()` function calculates the loss from stems which
died within a census interval. The arguments of the function are identical to
`obsSurGrowth()`:

```{r}
obsMorLoss(bicuar_clean, t0 = "2019", tT = "2021", w = "diam", 
  group = c("plot_id", "stem_id"), census = "census_date")
```

`obsRecGrowth()` calculates the growth from stems which recruited within
a census interval. 

There are multiple methods for estimating the growth of recruits. The default
`rec_method = "zero"`, assumes that the stem grew from a size of zero in the
initial census. While this method is very simple, it is often unrealistic in
instances where the minimum diameter threshold above which stems are measured
is greater than zero. 

```{r}
obsRecGrowth(bicuar_clean, "2019", "2021", w = "diam",
  group = c("plot_id", "stem_id"), census = "census_date",
  rec_method = "zero")
```

`rec_method = "thresh"` assumes that the stem grew from the minimum diameter
threshold of the plot. This method works well when growth is being calculated
in terms of diameter, as the value of `w_min_diam` can simply be a numeric
value defining the minimum diameter threshold of the plot. 

```{r}
obsRecGrowth(bicuar_clean, "2019", "2021", w = "diam",
  group = c("plot_id", "stem_id"), census = "census_date",
  rec_method = "thresh", w_min_diam = 5)
```

When `w` is something other than diameter, biomass for instance, this method
requires extra work. In this case, it is necessary to estimate what the biomass
of each stem would be if it shrunk down to the diameter of the minimum diameter
threshold of the plot. In the example below, biomass is estimated using a
simplistic allometric equation.

```{r}
# Define minimum diameter 
mdt <- 5

# Estimate AGB at minimum diameter 
bicuar_clean$agb_min <- (0.0644 * (pi * (mdt / 2) ^ 2) ^ 1.3341) / 1000

# Estimate growth from recruits
obsRecGrowth(bicuar_clean, "2019", "2021", w = "diam",
  group = c("plot_id", "stem_id"), census = "census_date",
  rec_method = "thresh", w_min_diam = "agb_min")
```

Finally, `rec_method = "extrap"` uses the average growth rate of the minimum
diameter size class to extrapolate the size of the recruit back to its
estimated size in the initial census. This method relies on having enough data
of sufficient quality to calculate an accurate 'average' growth rate for small
stems. `growth_percentile` refers to the percentile of growth rates used to
calculate the average growth rate of small stems. The default of 0.86 is
recommended by [@Talbot2014].

```{r}
obsRecGrowth(bicuar_clean, "2019", "2021", w = "diam",
  group = c("plot_id", "stem_id"), census = "census_date",
  rec_method = "extrap", diam = "diam", min_size_class = c(5, 10),
  min_diam_thresh = 5, growth_percentile = 0.86)
```

# Average growth rates

Rather than estimating individual stem growth rates separately for each census
interval, as in `obsSurGrowth()`, you may want to calculate the average growth
rate for each individual across all available census intervals. `growthMod()`
fits a linear model to each stem measurement timeline and extracts the model
slope coefficient, which is the average stem growth rate, in units of `w` per
`census`. Like `obsSurGrowth()`, with option `full = TRUE`, `growthMod()`
returns a dataframe with the columns in `group`, rather than a named vector.

```{r}
growthMod(bicuar_clean, w = "diam", group = c("plot_id", "stem_id"), 
  census = "census_date")
```

With the `growthAll()` function, you can also calculate growth increment for
every available consecutive or pairwise census interval for each stem. Where a
stem measurement in a particular census is missing, the function simply
calculates the growth rate using the next non-missing value:

```{r}
growthAll(bicuar_clean, w = "diam", group = c("plot_id", "stem_id"), 
  census = "census_date", type = "pairwise")
```

```{r}
(g_all <- growthAll(bicuar_clean, w = "diam", group = c("plot_id", "stem_id"), 
  census = "census_date", type = "consecutive"))
```

To convert these growth increments to a rate just divide by the census interval (`int`):

```{r}
g_all$g / g_all$int
```


# Areal growth rates

Measuring stem growth per unit area is best accomplished using the basal area
of all stems within the plot. First, calculate the basal area of each stem
measurement in units of m^2^, then aggregate to the total basal area within
each plot, and finally use the `growthAll()` function to calculate the basal
area growth for each census interval.

```{r}
bicuar_plot_ba <- bicuar_clean %>%
  mutate(ba = (pi * (diam/2)^2) / 10000) %>% 
  group_by(plot_id, census_date) %>% 
  summarise(ba_sum = sum(ba, na.rm = TRUE))

growthAll(bicuar_plot_ba, w = "ba_sum", group = "plot_id", 
  census = "census_date", type = "consecutive")
```

# Estimating unseen growth 

To accurately estimate productivity within a given census interval, i.e. the
total amount of biomass produced, as well as the observed growth of survivors
and the observed growth of recruits it is important to factor in the growth
from stems which died some time between the two censuses. [@Talbot2014]
provides a method for estimating this growth, assuming that the stems die
halfway through the census interval, and that they grow at the median growth
rate of their diameter size class.

```{r}
unobsMorGrowth(bicuar_clean, "2019", "2021", w = "diam", 
  group = c("plot_id", "stem_id"), census = "census_date", diam = "diam")
```

Additionally, there may be some stems which both recruit and then subsequently
die within a census interval, meaning they are never measured. Again,
[@Talbot2014] provides a method to both estimate the number of unseen recruits
which died (`unobsRec()`), and estimate their growth (`unobsRecGrowth()`). This
function only provides a single value, so we must subset the data to a single
plot and single census interval.

```{r}
bicuar_p1 <- bicuar_clean %>% 
  filter(plot_id == "ABG_5")

unobsRec(bicuar_p1, "2019", "2021", group = "stem_id", census = "census_date")
```

```{r}
unobsRecGrowth(bicuar_p1, "2019", "2021", w = "diam", group = "stem_id", 
  census = "census_date", diam = "diam", 
  min_size_class = c(5, 10), w_min_diam = 5)
```

Like `obsRecGrowth()` with the argument `rec_method = "extrap"`,
`unobsRecGrowth()` uses the median growth rate of the smallest diameter size
class to estimate the growth rate of the unseen recruits. The function assumes
that these stems recruit in about 1/3 of the way through the census interval,
and die about 2/3 of the way through.

# Plot-level rates of biomass change, productivity, loss

The functions above present parts of a workflow that can be used to generate
estimates of plot-level productivity, biomass turnover, and biomass change.
There are two functions in `{ProdVital}` which provide these plot-level
statistics. `prodKohyama()` uses methods from [@Kohyama2019], while
`prodTalbot()` uses methods from [@Talbot2014]. 

`prodTalbot()` uses `unobsRecGrowth()` and `unobsMorGrowth()` to estimate the
unseen productivity from stems which died, while `prodKohyama()` simply uses
the observed initial and final biomass values of survivors and the population,
and accounts for unseen productivity using estimates of population level
processes.

```{r}
prodKohyama(bicuar_p1, "2019", "2021", w = "diam",
  group = "stem_id", census = "census_date")
```

```{r}
prodTalbot(bicuar_p1, "2019", "2021", w = "agb", diam = "diam",
  group = "stem_id", census = "census_date", w_min_diam = "agb_min", 
  rec_method = "zero")
```

# Plot-level vital rates

Estimates of population demographic processes are equally important to
understanding ecosystem function as estimates of productivity. `{ProdVital}`
provides `vitalKohyama()` to estimate rates of recruitment, mortality and other
derived metrics, using methods from [@Kohyama2018]. 

```{r}
vitalKohyama(bicuar_p1, "2019", "2021", group = "stem_id", 
  census = "census_date", plot_area = 1)
```

# Scaling up to multiple plots and multiple census intervals

All the functions in `{ProdVital}` act on a single plot across a single census
interval, but it is trivial to run the functions across groups of plots and
census intervals using `lapply()` or `for` loops.

First, split the dataset into plots:

```{r}
bicuar_split <- split(bicuar_clean, bicuar_clean$plot_id)
```

Then, for each plot, identify all the possible pairwise combinations of
censuses:

```{r}
prod_all <- do.call(rbind, lapply(bicuar_split, function(x) { 
  # Find all census dates
  census_date_all <- sort(unique(x$census_date))

  # Create all pairwise combinations of censuses
  comb_list_pair <- combn(census_date_all, 2, simplify = FALSE)

  pK_census <- lapply(comb_list_pair, function(y) { 
    # Estimate productivity according to Kohyama 
    prodKohyama(x, t0 = y[1], tT = y[2], w = "agb", 
      group = "stem_id", census = "census_date")
  })

  pT_census <- lapply(comb_list_pair, function(y) { 
    # Estimate productivity according to Kohyama 
    prodTalbot(x, t0 = y[1], tT = y[2], w = "agb", diam = "diam",
      group = "stem_id", census = "census_date", w_min_diam = "agb_min", 
      rec_method = "zero")
  })

  # Create dataframes of metrics
  pK <- data.frame()[seq_along(pK_census), ]
  for (y in names(pK_census[[1]])) {
    pK[[y]] <- unlist(lapply(pK_census, "[[", y))
  }

  pT <- data.frame()[seq_along(pT_census), ]
  for (y in names(pT_census[[1]])) {
    pT[[y]] <- unlist(lapply(pT_census, "[[", y))
  }

  # Join the dataframes
  out <- left_join(pK, pT, by = c("t0", "tT", "int"))

  # Add plot ID
  out$plot_id <- unique(x$plot_id)

  out
}))

prod_all
```

# Comparing Kohyama and Talbot

Some of the metrics returned by `prodKohyama()` and `prodTalbot()` are
analogous, though their values may differ due to the different methods used by
each function.

For example, comparing annual rates of productivity `P_ann` and `AGWP_est_ann`:

```{r}
ggplot(prod_all, aes(x = P_ann, y = AGWP_obs_ann)) + 
  geom_abline(linetype = 2, colour = "red") +
  geom_point() 
```

# References

